Effects of PCSK9 Inhibition With Evolocumab on Lipoprotein Metabolism in subjects with elevated Lp(a)

### 2019-10-09 manuscript.Rmd
introduction for draft version 3

Lipoprotein(a) [Lp(a)] is an independent risk factor for cardiovascular diseases, and often considered as a low-density lipoprotein (LDL)-like particle with an apolipoprotein(a) [apo(a)] tail produced by *LPA*. 
Unlike LDL-cholesterol lowering, limited therapeutic options are available to reduce plasma Lp(a) concentration. 
Proprotein convertase subtilisin/kexin type 9 (PCSK9) monoclonal antibody such as evolocumab has been shown to lower Lp(a) [@Raal2014; @Watts2017; @Shapiro2019].
PCSK9 binds to hepatic low density lipoprotein (LDL) receptors and mediates their degradation in lysosome.
Evolocumab (also called AMG 145) is a fully human monoclonal antibody that targets PCSK9 and prevents it from binding to the LDL receptor. 
Although evolocumab-induced Lp(a) reduction was linked to enhanced LDL receptor mediated uptake and attenuated Lp(a) production, the mechanism of Lp(a) lowering by PCSK9 inhibition is unknown [@Romagnuolo2015; @Raal2016; @Watts2018]. 
As shown in a recent meta-analysis, usage of statins leads to increasing of Lp(a) concentration [@Tsimikas2019]. 
More interestingly, decreasing Lp(a) concentration was seen in homozygous familial hypercholesterolaemia (FH) patients after treatment of evolocumab [@Raal2015]. 
These observations suggested existence of non-LDL receptor mediated catabolism of Lp(a). 
In the current study, we aim to explore the alternative pathways beyond LDL receptor-mediated catabolism of Lp(a).
To this end, we studied the metabolic effects of PCSK9 inhibition with evolocumab in 30 subjects (placebo 14; evolocumab 16) with elevated Lp(a). 
We identified that despite evolocumab treatment for 16 weeks resulted in 18.9% reduction of Lp(a), coupled with substantial reduction of VLDL, IDL and LDL particles.   
Our further analysis revealed that evolocumab treatment not only reduced LDL particle concentrations, but also changed the distribution between LDL particles with and without apo(a).

>>>>

lipoprotein(a) [Lp(a)], synthesized in liver, is a low-density lipoprotein (LDL)-like particle with an apolipoprotein(a) [apo(a)] tail. 
Similar to other lipoproteins, Lp(a) is cleared from the circulation primarily by the liver [@Cain2005]. 
Epidemiological studies showed that subjects with elevated Lp(a) had increasing risk of developing cardiovascular disease [@ODonoghue2019]. 
Unlike LDL-cholesterol lowering, limited therapeutic options are available to reduce plasma Lp(a) concentration [@VanCapelleveen2016]. 
One of the agents that are capable to lower circulating Lp(a) levels is proprotein convertase subtilisin/kexin type 9 (PCSK9) monoclonal antibody, which prevents extracellular PCSK9 from binding to the LDL receptor, and results in supraphysiological abundance of LDL receptors [@Raal2014; @Raal2016].
The excess number of LDL receptors on the surface of hepatocytes were hypothesized to be responsible for the enhanced Lp(a) catabolism [@Romagnuolo2015; @Raal2016]. 
However, the above hypothesis does not concur with the effect of statins therapy, which significantly increases hepatic LDL receptor abundances but also increases plasma Lp(a) levels [@Tsimikas2019]. 
The same hypothesis was also challenged by the observation that LDL receptor knock out mice had similar Lp(a) catabolism as the wild type mice [@Cain2005]. 
Therefore, what roles LDL-receptor plays in Lp(a) catabolism has not been elucidated and remains a matter of debate [@Reyes-Soffer2017; @Watts2018; @Stoekenbroek2018]. 
In addition to the LDL receptor, very low density lipoprotein (VLDL) receptor and scavenger receptor class B type I (SR-BI) were also found to act as potential receptor for Lp(a), indicating more than one pathway is likely to be involved in Lp(a) reduction by PCKS9 inhibition [@Argraves1997; @Yang2013].
As a cost-effective approach, nuclear magnetic resonance (NMR) metabolomics simultaneously quantifies hundreds of circulating biomarkers and provides insights on multiple pathways [@Soininen2009].
In fact, NMR metabolomics has been used to study metabolic effects that were associated with  *LPA* loci, as well as genetic inhibition of PCSK9 [@Kettunen2016; @Sliz2018]. 
However, genetic inhibition of PCSK9 is different from inhibition with monoclonal antibody, and there is lack of studies on metabolic effects of PCSK9 inhibition with monoclonal antibody. 
In the current study, we aim to study metabolic changes corresponding to Lp(a) reduction in subjects with elevated Lp(a) after treatment of PCSK9 monoclonal antibody (evolocumab).
To this end, we performed NMR metabolomics of plasma samples derived from 30 healthy participants with elevated Lp(a) levels (placebo 14; evolocumab 16). 
Since Lp(a) and LDL significantly overlap in density, protein and lipid composition, LDL measurements made by current technologies contain the content of Lp(a) [@Yeang2015]. 
To address this issue, in this study we developed a statistical model to distinguish these two lipoproteins.
